---
title: "bellabeat_EDA"
author: "Deni"
date: "1/30/2022"
output: html_document
---

### About BellaBeat
  * A high-tech company that manufactures health-focused smart products.
  * BellaBeat collects data on activity, sleep, stress, and reproductive health to empower women with knowledge about their own health and habits.
  * Founded in 2013, Bellabeat has grown rapidly and quickly positioned itself as a tech-driven wellness company for woman.

##### Stakeholders
  * Urška Sršen: Bellabeat’s cofounder and Chief Creative Officer
  * Sando Mur: Mathematician and Bellabeat’s cofounder; key member of the Bellabeat executive team
  * Bellabeat marketing analytics team

#### Business Task for this Case Study:
  * Analyze smart device usage data in order gain insight into how consumers use non-Bellabeat smart devices.
  * Then select one product to apply these insights to.

#### Questions to answer:
  * What are some trends that can be found in smart device usage?
  * How could we apply these trends to Bellabeat customers?
  * How could these trends influence Bellabeats marketing strategy?
  
Data set used for this EDA https://www.kaggle.com/arashnic/fitbit.

First we will be installing and loading the required R packages.
```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
```
## Dive into the datasets
```{r}
daily_Activity_merged <- read_csv('dailyActivity_merged.csv')
daily_Calories_merged <- read_csv('dailyCalories_merged.csv')
daily_Steps_merged <- read_csv('dailySteps_merged.csv')
daily_sleep_merged <- read_csv('sleepDay_merged.csv')
```
#### Get an Idea of what data is kept inside each set.
```{r}
head(daily_Activity_merged)
```
```{r}
head(daily_Calories_merged)
```
```{r}
head(daily_Steps_merged)
```

```{r}
head(daily_sleep_merged)
```
We can add a new column to daily_sleep_merged where we can see how much time is spent in bed but not asleep by creating a new column named Time_Spent_Awake_in_Bed by subtracting TotalTimeInBed by TotalMinutesAsleep.

```{r}
daily_sleep_merged <- daily_sleep_merged %>% mutate(Time_Spent_Awake_In_Bed = daily_sleep_merged$TotalTimeInBed - daily_sleep_merged$TotalMinutesAsleep)
head(daily_sleep_merged)
```
We can also see that the SleepDay column contains the time as well. We should seperate each day and time into two different columns.
```{r}
daily_sleep_merged <- daily_sleep_merged %>% separate(SleepDay, c("Date", "Time"), " ")
head(daily_sleep_merged)
```



##### Find common columns between each of the datasets.
```{r}
colnames(daily_Activity_merged)
```
```{r}
colnames(daily_Calories_merged)
```
```{r}
colnames(daily_sleep_merged)
```
```{r}
colnames(daily_Steps_merged)
```
We see a common trait between each data set and that each of them contain an "ID" column. Now I will check to see how many unique sets of ID's there are.
```{r}
n_distinct(daily_Activity_merged$Id)
```

```{r}
n_distinct(daily_Calories_merged$Id)
```
```{r}
n_distinct(daily_sleep_merged$Id)
```
```{r}
n_distinct(daily_Steps_merged$Id)
```
We see that we have 33 Unique Id's from DailyActivity, DailyCalories, and DailySteps datasets.
While we only have 24 Unique Id's from the DailySleep dataset.

With an unequal amount of Id's I will check to see if there are any duplicates in the DailyActivity, DailyCalories, and DailySteps datasets.
```{r}
nrow(daily_Activity_merged[duplicated(daily_Activity_merged),])
```
```{r}
nrow(daily_Calories_merged[duplicated(daily_Calories_merged),])
```
```{r}
nrow(daily_Steps_merged[duplicated(daily_Steps_merged),])
```
We find a result of no duplicates in each dataset. I will now check if DailySleep has any duplicates
```{r}
nrow(daily_sleep_merged[duplicated(daily_sleep_merged),])
```
We find 3 duplicated rows in the dataset. I will now create a new dataframe without the 3 duplicated rows.
```{r}
daily_sleep_merged_cleaned <- unique(daily_sleep_merged)
nrow(daily_sleep_merged_cleaned[duplicated(daily_sleep_merged_cleaned),])
```
### Now we Analyze the data
  * By gathering statistical data from each dataset.

```{r}
head(daily_Activity_merged)
#we want total steps, totaldistance, VeryActiveMinutes, LightlyActiveMinutes, SedentaryMinutes, and Calories.
daily_Activity_merged %>% 
  select(TotalSteps, TotalDistance, VeryActiveMinutes, LightlyActiveMinutes, SedentaryMinutes, Calories) %>% 
    summary()
```
#### Daily Activity Statistics
  * We see the Maximum amount of steps performed by a person was 36,019.
  * We also get to see that the Maximum amount of VeryActiveMinutes are 210 minutes while the Maximum amount of SedentaryMinutes are 1440 Minutes
  * The average amount of steps performed is 7,406.
  * As well as the average amount of VeryActiveMinutes is 21 compared to the average amount of SedentaryMinutes are 991.
  
Next we take a look at the statistics from the daily calorie set.
```{r}
head(daily_Calories_merged)
summary(daily_Calories_merged)
```
#### Daily Calories Statistics
  * We see that the maximum amount of calories gained from the 33 participants are 4,900 calories gained in a day.
  * While the average is 2,304.
  * With an average of 2,000 calories needed daily for the average human we can say that the average person in this dataset are achieving their daily calories.
  
Now onto the daily steps Statistics
```{r}
head(daily_Steps_merged)
summary(daily_Steps_merged)
```
#### Daily Steps Statistics
  * We see a Maximum of 36,019 steps gathered in one day. We can conclulde that, the person who achieved that many steps in a day is very active.
  * The average steps in a day is 7,638.
  

Now onto the daily sleep dataset.
```{r}
head(daily_sleep_merged_cleaned)
daily_sleep_merged_cleaned %>% select(Id, TotalSleepRecords, TotalMinutesAsleep, TotalTimeInBed, Time_Spent_Awake_In_Bed) %>% summary()
```
#### Daily Sleep Statistics
  * We gathered that the most amount of sleep gained by an individual was 796 minutes which converts to a little over 13 hours!
  * While the average time spent asleep is 419 minutes which converts to 7 hours.
  * We gathered that the most time spent awake in bed is 371 minutes which converts to 6 hours.
  * While the average time spent awake in bed is 39 minutes.

  
## Onto Visualizing the Data.
  * First we will visualize the dailyActivity dataset to showcase how many calories each person had burned compared to how many steps they took.
```{r}
ggplot(data = daily_Activity_merged) + geom_jitter(mapping = aes(x = TotalSteps, y = Calories), color = 'blue') +
  geom_smooth(mapping = aes(x = TotalSteps, y = Calories), color = "red") +
  labs(title = "Total Steps Taken VS Total Calories Burned in a Day", x = "Total Steps", Y = "Calories")
```

The graph above shows the correlation between the total amount of steps taken compared to the total amount of calories burned.
We can gather the fact that the more steps that are taken that the more calories are burned.

Now lets check the relationship between sedentary minutes and calories burned.
```{r}
ggplot(data = daily_Activity_merged) + geom_jitter(mapping = aes(x = SedentaryMinutes, y = Calories), color = "Green") +
  geom_smooth(mapping = aes(x = SedentaryMinutes, y = Calories), color = "Red") + 
  labs(title = "The Relationship between Minutes Inactive and Calories Burned", x = "Inactive Minutes", y = "Calories Burned")
```

The graph above shows that the less sedentary minutes a person has throughout the day the more calories they burn.
  
Now lets compare Time Spent awake in Bed and Total Minutes Asleep
```{r}
ggplot(data = daily_sleep_merged_cleaned) + geom_jitter(mapping = aes(x = Time_Spent_Awake_In_Bed, y = TotalMinutesAsleep), color = 'green') +
  geom_smooth(mapping = aes(x = Time_Spent_Awake_In_Bed, y = TotalMinutesAsleep), color = 'red') +
  labs(title = "The Relationship Between Time Spent Not Sleeping While in Bed and Total Minutes Asleep", x = 'Minutes Awake', y = 'Minutes Asleep')

```

From the graph above we can gather that the less minutes spent awake in bed the more sleep the average person receives.

### Changes to BellaBeat's marketing strategy based off of the trend we found in the DataSets.
  * Based off the results received from the data, users burn more calories the more minutes they're active. So, Bellabeat should showcase this data to their users to be more active through notifications. 
  * As well as showcase this data to their users who are more active by congratulating them when they have had daily active streaks.
  * And allow users to set daily goals for total steps taken in a day or total calories burned within a day.
  


